The Unicode Cookbook \linebreak for Linguists

### 2018-06-08 book/chapters/orthography_profiles.tex
deals with all paperhiver comments except a few time consuming ones

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode normalization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA}). 
Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated canonically equivalent for search and comparison.

>>>>

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode normalization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA}). 
Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated  for search and comparison.

### 2018-06-08 book/chapters/pitfalls.tex
deals with all paperhiver comments except a few time consuming ones

The character, represented here by glyph <Å>, is encoded in the Unicode Standard
in the first two examples by a single-character sequence; each is assigned a
different code point. In the third example, the glyph is encoded in a
multiple-character sequence that is composed of two character code points. All
three sequences are \textsc{canonically equivalent}, i.e.~they are strings that
represent the same abstract character and because they are not distinguishable
by the user, the Unicode Standard requires them to be treated the same in
regards to their behavior and appearance. Nevertheless, they are encoded
differently. For example, if one were to search an electronic text (with
software that does not apply Unicode Standard normalization) for
\textsc{angstrom sign} (\uni{212B}), then the instances of \textsc{latin 
capital letter a with ring above} (\uni{00C5}) would not be found.

@@

As illustrated, some characters in the Unicode Standard have alternative
representations (in fact, many do), but the Unicode Normalization Algorithm can
be used to transform certain sequences of characters into canonical
forms to test for equivalency. To determine equivalence, each
character in the Unicode Standard is associated with a combining class, which is
formally defined as a character property called \textsc{canonical combining
class} which is specified in the Unicode Character Database. The combining class
assigned to each code point is a numeric value between 0 and 254 and is used by
the Unicode Canonical Ordering Algorithm to determine which sequences of
characters are canonically equivalent. Normalization forms, as very briefly
described above, can be used to ensure character equivalence by ordering
character sequences so that they can be faithfully compared.

It is very important to note that any software application that is Unicode
Standard compliant is free to change the character stream from one
representation to another. This means that a software application may compose,
decompose or reorder characters as its developers desire; as long as the
resultant strings are canonically equivalent to the original. This might lead to
unexpected behavior for users. Various players, like the Unicode Consortium, the
W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and
some software applications that we tested indeed seem to automatically convert
strings into NFC.\footnote{See the summary of various recommendation here:
\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.} This means in
practice that if a user, for example, enters <a> and <\dia{0300}>, i.e.~\textsc{latin
small letter a} at \uni{0061} and \textsc{combining grave accent} at \uni{0300},
this might be automatically converted into <à>, i.e.~\textsc{latin small letter
a with grave} at \uni{00E0}.\footnote{The behavior of software applications can
be quite erratic in this respect. For example, Apple's TextEdit does not do any
conversion on text entry. However, when you copy and paste some text inside the
same document in rich text mode (i.e.~RTF-format), it will be transformed into
NFC on paste. Saving a document does not do any conversion to the glyphs on
screen, but it will save the characters in NFC.}

@@

There are further combinations that deserve special care because it is actually
possible to produce identical characters in different ways without them being
canonically equivalent. In these situations, the general rule holds, namely that
characters with attached extras are not decomposed. However, in the following
cases the extras actually exist as combining diacritics, so there is also 
the possibility to construct a character by using a base character with those 
combining diacritics.

>>>>

The character, represented here by glyph <Å>, is encoded in the Unicode Standard
in the first two examples by a single-character sequence; each is assigned a
different code point. In the third example, the glyph is encoded in a
multiple-character sequence that is composed of two character code points. All
three sequences are \textsc{}, i.e.~they are strings that
represent the same abstract character and because they are not distinguishable
by the user, the Unicode Standard requires them to be treated the same in
regards to their behavior and appearance. Nevertheless, they are encoded
differently. For example, if one were to search an electronic text (with
software that does not apply Unicode Standard normalization) for
\textsc{angstrom sign} (\uni{212B}), then the instances of \textsc{latin 
capital letter a with ring above} (\uni{00C5}) would not be found.

@@

As illustrated, some characters in the Unicode Standard have alternative
representations (in fact, many do), but the Unicode Normalization Algorithm can
be used to transform certain sequences of characters into canonical
forms to test for equivalency. To determine equivalence, each
character in the Unicode Standard is associated with a combining class, which is
formally defined as a character property called \textsc{canonical combining
class} which is specified in the Unicode Character Database. The combining class
assigned to each code point is a numeric value between 0 and 254 and is used by
the Unicode Canonical Ordering Algorithm to determine which sequences of
characters are . Normalization forms, as very briefly
described above, can be used to ensure character equivalence by ordering
character sequences so that they can be faithfully compared.

It is very important to note that any software application that is Unicode
Standard compliant is free to change the character stream from one
representation to another. This means that a software application may compose,
decompose or reorder characters as its developers desire; as long as the
resultant strings are  to the original. This might lead to
unexpected behavior for users. Various players, like the Unicode Consortium, the
W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and
some software applications that we tested indeed seem to automatically convert
strings into NFC.\footnote{See the summary of various recommendation here:
\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.} This means in
practice that if a user, for example, enters <a> and <\dia{0300}>, i.e.~\textsc{latin
small letter a} at \uni{0061} and \textsc{combining grave accent} at \uni{0300},
this might be automatically converted into <à>, i.e.~\textsc{latin small letter
a with grave} at \uni{00E0}.\footnote{The behavior of software applications can
be quite erratic in this respect. For example, Apple's TextEdit does not do any
conversion on text entry. However, when you copy and paste some text inside the
same document in rich text mode (i.e.~RTF-format), it will be transformed into
NFC on paste. Saving a document does not do any conversion to the glyphs on
screen, but it will save the characters in NFC.}

@@

There are further combinations that deserve special care because it is actually
possible to produce identical characters in different ways without them being
. In these situations, the general rule holds, namely that
characters with attached extras are not decomposed. However, in the following
cases the extras actually exist as combining diacritics, so there is also 
the possibility to construct a character by using a base character with those 
combining diacritics.

### 2018-06-08 book/chapters/practical_recommendations.tex
Resolves issue #29
deals with all paperhiver comments except a few time consuming ones

This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in Chapters \ref{orthography-profiles} and \ref{implementation}. Instead, we provide links to quickly find general information about the Unicode Standard and the International Phonetic Alphabet (IPA). And we target ordinary working linguists who want to know how to easily insert special characters into their digital documents and applications.

@@

There are also Web-based Unicode character pickers available through the browser that allow for the creation and insert of special characters, which can then be copied \& pasted into documents or software applications. For example, try:

@@

On Windows, accented characters can be inserted by using alt-key shortcuts, i.e.\ holding down the alt-key and keying in a sequence of numbers (which typically reflect the Unicode character's decimal representation). For example, \textsc{latin small letter c with cedilla} at \uni{00E7} with the decimal code 231 can be inserted by holding the alt-key and keying the sequence 0231. Again, this method is burdensome for rapidly inserting characters. For access to the full range of Unicode characters, the Character Map program comes preinstalled on all Microsoft Operating systems.

There are also many third party applications that provide virtual keyboards. These programs typically override keys or keystrokes on the user's keyboard  allowing them to quickly keyboard special characters (once the layout of the new keyboard is mastered). They can be language-specific or devoted specifically to IPA. Two popular programs are:

@@

There are several good Unicode IPA character pickers available on the Web and through the browser, including:

@@

Regarding fonts that display Unicode IPA correctly, many linguists turn to the IPA Unicode fonts developed by SIL Internnational. The complete SIL font list is available online.\footnote{\url{http://scripts.sil.org/SILFontList}} There is also a page that describes IPA transcription using the SIL fonts and provides an informative discussion on deciding which font to use.\footnote{\url{http://scripts.sil.org/ipahome}} Traditionally, IPA fonts popular with linguists were created and maintained by SIL International, so it is often the case in our experience that we encounter linguistics data in legacy IPA fonts, i.e.\ pre-Unicode fonts such as SIL IPA93.\footnote{\url{http://scripts.sil.org/FontFAQ_IPA93}} SIL International does a good job of describing how to convert from legacy IPA fonts to Unicode IPA. The most popular Unicode IPA fonts are Doulos SIL and Charis SIL:

@@

For users of Python, see the standard documentation on how to use Unicode in your programming applications.\footnote{\url{https://docs.python.org/3/howto/unicode.html}} For R users we recommend the \textsc{stringi} library.\footnote{\url{https://cran.r-project.org/web/packages/stringi/index.html}} For \LaTeX users, the TIPA package is useful for inserting IPA characters into your typeset documents. See these resources:

@@

\noindent But we find it much easier to use the Unicode-aware XeTeX/XeLaTeX typesetting system.\footnote{\url{http://xetex.sourceforge.net/}} Unicode characters can be directly inserted into your Tex documents and compiled into typeset PDF with XeLaTeX.

>>>>

This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in Chapters \ref{orthography-profiles} and \ref{implementation}. Instead, we provide links to quickly find general information about the Unicode Standard and the International Phonetic Alphabet (IPA). We target ordinary working linguists who want to know how to easily insert special characters into their digital documents and applications.

@@

There are also web-based Unicode character pickers available through the browser that allow for the creation and insertion of special characters, which can then be copied \& pasted into documents or software applications. For example, try:

@@

On Windows, accented characters can be inserted by using alt-key shortcuts, i.e.\ holding down the alt-key and keying in a sequence of numbers (which typically reflect the Unicode character's decimal representation). For example, \textsc{latin small letter c with cedilla} at \uni{00E7} with the decimal code 231 can be inserted by holding the alt-key and keying the sequence 0231. Again, this method is burdensome for rapidly inserting characters. For access to the full range of Unicode characters, the Character Map program comes preinstalled on all Microsoft operating systems.

There are also many third-party applications that provide custom keyboard layouts. These programs typically override keys or keystrokes on the user's keyboard allowing them to quickly enter special characters (once the layout of the new keyboard is mastered). They can be language-specific or devoted specifically to IPA. Two popular programs are:

@@

There are several good Unicode IPA character pickers available through the browser, including:

@@

Regarding fonts that display Unicode IPA correctly, many linguists turn to the IPA Unicode fonts developed by SIL International. The complete SIL font list is available online.\footnote{\url{http://scripts.sil.org/SILFontList}} There is also a page that describes IPA transcription using the SIL fonts and provides an informative discussion on deciding which font to use.\footnote{\url{http://scripts.sil.org/ipahome}} Traditionally, IPA fonts popular with linguists were created and maintained by SIL International, so it is often the case in our experience that we encounter linguistics data in legacy IPA fonts, i.e.\ pre-Unicode fonts such as SIL IPA93.\footnote{\url{http://scripts.sil.org/FontFAQ_IPA93}} SIL International does a good job of describing how to convert from legacy IPA fonts to Unicode IPA. The most popular Unicode IPA fonts are Doulos SIL and Charis SIL:

@@

For users of Python, see the standard documentation on how to use Unicode in your programming applications.\footnote{\url{https://docs.python.org/3/howto/unicode.html}} For R users we recommend the \textsc{stringi} library.\footnote{\url{https://cran.r-project.org/web/packages/stringi/index.html}} For \LaTeX~users the \textsc{TIPA} package is useful for inserting IPA characters into your typeset documents. See these resources:

@@

\noindent But we find it much easier to use the Unicode-aware \hologo{XeTeX} typesetting system.\footnote{\url{http://xetex.sourceforge.net/}} Unicode characters can be directly inserted into your \hologo{TeX} documents and compiled into typeset PDF with \hologo{XeLaTeX}.

### 2018-06-08 book/chapters/unicode_approach.tex
deals with all paperhiver comments except a few time consuming ones

Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
version 11) consists of a set of specifications and guidelines that include (i) a
core specification, (ii) code charts, (iii) standard annexes and (iv) a
character database.\footnote{All documents of the Unicode Standard are available
at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
of terminology inside the Unicode Standard, their glossary is particularly
useful, available at \url{http://www.unicode.org/glossary/}. For a general
introduction to the principles of Unicode, Chapter 2 of the core specification,
called \textsc{general structure}, is particularly insightful. Unlike 
many other documents in the Unicode Standard, this general introduction is
relatively easy to read and illustrated with many interesting examples from
various orthographic traditions from all over the world.} The \textsc{core
specification} is a book aimed at human readers that describes the formal
standard for encoding multilingual text. The \textsc{code charts} provide a
human-readable online reference to the character contents of the Unicode
Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
are a set of technical standards that describe the implementation of the Unicode
Standard for software development, Web standards, and programming languages.
The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
files that describe the character properties, including a set of rich character
and writing system semantics, for each character in the Unicode Standard. In
this section, we introduce the basic Unicode concepts, but we will leave out
many details. Please consult the above-mentioned full documentation for a more
detailed discussion. Further note that the Unicode Standard is exactly that,
namely a standard. It normatively describes notions and rules to be followed. In
the actual practice of applying this standard in a computational setting, a
specific implementation is necessary. The most widely used implementation of the
Unicode Standard is the \textsc{International Components for Unicode (ICU)},
which offers C/C++ and Java libraries implementing the Unicode
Standard.\footnote{More information about the ICU is available here:
\url{http://icu-project.org}.}

>>>>

Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
version 11) consists of a set of specifications and guidelines that include (i) a
core specification, (ii) code charts, (iii) standard annexes and (iv) a
character database.\footnote{All documents of the Unicode Standard are available
at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
of terminology inside the Unicode Standard, their glossary is particularly
useful, available at \url{http://www.unicode.org/glossary/}. For a general
introduction to the principles of Unicode, Chapter 2 of the core specification,
called \textsc{general structure}, is particularly insightful. Unlike 
many other documents in the Unicode Standard, this general introduction is
relatively easy to read and illustrated with many interesting examples from
various orthographic traditions from all over the world.} The \textsc{core
specification} is a book aimed at human readers that describes the formal
standard for encoding multilingual text. The \textsc{code charts} provide a
human-readable online reference to the character contents of the Unicode
Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
are a set of technical standards that describe the implementation of the Unicode
Standard for software development, web standards, and programming languages.
The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
files that describe the character properties, including a set of rich character
and writing system semantics, for each character in the Unicode Standard. In
this section, we introduce the basic Unicode concepts, but we will leave out
many details. Please consult the above-mentioned full documentation for a more
detailed discussion. Further note that the Unicode Standard is exactly that,
namely a standard. It normatively describes notions and rules to be followed. In
the actual practice of applying this standard in a computational setting, a
specific implementation is necessary. The most widely used implementation of the
Unicode Standard is the \textsc{International Components for Unicode (ICU)},
which offers C/C++ and Java libraries implementing the Unicode
Standard.\footnote{More information about the ICU is available here:
\url{http://icu-project.org}.}
